Numerical modelling of the Cordilleran ice sheet

### 2014-07-19 kappa.tex
Rework 'perspectives' into 'sources of error'.

% ======================================================================
\section{Conclusion and perpectives}
% ======================================================================

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, not been attained
before. The coupled SIA-SSA approximation is crude, yet it is currently the
only approach that allows, combined with massively parallel computing, high
spatial resolution over the multi-millenial time scales of ice sheet growth and
decay.

Through sensitivity studies, I have identified the predominant sources of
uncertainty, which I have assessed using a combination of detailed
meteorological datasets and available geological evidence.\todo{insert a
statement which these sources of uncertainties are} Simulations supporting this
thesis yielded estimates of the LGM Cordilleran ice sheet extent and volume,
identification of the main ice-dispersal centres througout the last glacial
cycle, and understanding of final deglaciation patterns. However, there remains
considerable room for model improvements and further analysis of the simulation
output \todo{introduce your perspectives here}.

% ----------------------------------------------------------------------
\subsection{Precipitation change}
% ----------------------------------------------------------------------

One of the predominant uncertainties, for which Cordilleran ice sheet modelling
is highly sensitive, is the treatment of precipitation. Throughout the
modelling, we base precipitation patterns on current patterns derived from
reanalysis data. Although the distribution of precipitation has undoubtedly
been affected by past changes in global climate and the intermittent presence
of ice sheets and ice caps in the Northern Hemisphere, its changes are
difficult to assess. In the North American Cordillera, much of the continental
interior was likely dryer than today due to the presence of ice caps on the
Pacific Coast ranges, reinforcing the current topographic barrier. However,
some regions were probably rainier as a side effect of modified wind patterns.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown in \CCLI\ that different atmospheric
reanalyses which are circulation models that use data assimilation techniques,
fail to agree regarding the present-day pattern of precipitation over the
modelling domain. These discrepancies, which arise as a consequence of the
highly mountainous terrain, have dramatic effects on the ice sheet model
response. Considering additional difficulties inherent to modelling past
climates, and the lack of data for assimilation, we must regard derived ice
sheet histories as a highly generalized model output [only expect that
palaeo-climate reconstructions will produce even higher discrepancies in the
ice-sheet-model response].

\todo{think tomorrow. can one say something about the robustness of some
      features (ice caps that form, spreading centers which persist)which occur
      irrespective of which reanalysis is used?}.

% ----------------------------------------------------------------------
\subsection{Geothermal heat flux}
% ----------------------------------------------------------------------

Geothermal heat flux, contributing to the en-glacial temperature distribution
and the location of basal melt areas, was assumed constant over the modelling
grid. However, the geology of the study area consists of an amalgamation of
former subduction zone, resulting in complex geothermal patterns
\citep{Blackwell.Richards.2004}. I am now at the stage to start including this
variability as a driver to the ice sheet model, which will offer new
opportunities to study its output in terms of ice flow patterns, consistent
basal freezing and basal melting patterns, regions prone for glacial erosion,
and regions characterised by glacial deposition.

I expect that inclusion of geothermal heat patterns allow modelled basal flow
directions and distances to be compared with glacial lineations, which is one
of the main sources of geomorphological evidence used to reconstruct former ice
sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
instance, in our simulations, large portions of the Cordilleran ice sheet
remain cold-based throughout deglaciation (\CCYC, Fig.~15). As cold-based
glaciers typically do not impact their beds significantly, this could indicate
that some of the lineation patterns we observe within these zones of consistent
cold-based deglaciation could be older features, formed during an earlier phase
of advance or retreat of the ice sheet. However, without a better control on
the geothermal heat flux, such analysis are perhaps too premature.

% ----------------------------------------------------------------------
\subsection{The Cassiar ice cap}
% ----------------------------------------------------------------------

In the summer of 2012, I have co-organised and participated in a fieldwork with
Martin Margold and Ping Fu to collect samples of glacial deposits in the
southern Cassiar Mountains of northern British Columbia
(\cref{fig:map-cordillera}). The primary target of the fieldwork was a
triple-crested, sharp, ca. 50\,m-high moraine located in the vicinity of Tuya
Lake on the western slope of the Cassiar Mountains
(\cref{fig:photo-tuya-moraine}). Although this is a unique feature in the area,
the size of the moraine and its location at the foothill of the mountain range
indicate that it was likely formed by a tributary glacier from an ice cap,
rather than a stand-alone valley glacier. The relatively low-elevated Cassiar
Mountains do not appear as a major nucleation centre in our simulations (\CCYC,
Fig.~5). Therefore, it is tempting to think that such a large feature may
result from the late-glacial re-advance of a remnant of the decaying ice sheet.
High-resolution modelling of the deglaciation phase, perhaps at a regional
scale, may help to study this hypothesis.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
  \caption{The Tuya Lake Moraine in northern British Columbia.}
  \label{fig:photo-tuya-moraine}
\end{figure}

>>>>

% ======================================================================
\section{Sources of error}
% ======================================================================

Through sensitivity studies, I have identified the predominant sources of
uncertainty involved in numerical modelling of Cordilleran ice sheet. Through
the different papers constituting this thesis, I present an assessment of some
of these sources, including the choice of present-day atmospheric forcing, the
effects of daily temperature variability, and the choice of palaeo-temperature
time series used to drive the model through past climate history.

The resulting simulations, combined with available geological evidence,
yielded estimates of the LGM Cordilleran ice sheet extent and volume,
identification of the main ice-dispersal centres througout the last glacial
cycle, and understanding of final deglaciation patterns. However, there remains
considerable room for model improvements and further analysis of the simulation
output. Sources of error not assessed in this thesis include potential
buttressing effects of the Laurentide ice sheet, past changes in precipitation,
and regional variability of the geothermal heat flux.

% ----------------------------------------------------------------------
\subsection{Interaction with the Laurentide ice sheet}
% ----------------------------------------------------------------------

Certainly the most important assumption made in this thesis is that of a
Cordilleran ice sheet largely disconnected from its much larger neighbour, the
Laurentide ice sheet. Geomorphological evidence from the suture zone indicates
that, during the LGM, both Cordilleran and Laurentide ice flow has been
deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
Margold.etal.2013a}, although it remains unclear for how long this junction has
lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007, Gowan.2013}.

Nevertheless, during most of the last glacial cycle, the two ice sheets have
been disconnected \citep{Kleman.etal.2010}. Thus, this assumption only affects
numerical reconstructions of the Cordilleran ice sheet presented here during
the last glacial maximum and during the subsequent deglacation. Other numerical
modelling results \citep{Gregoire.etal.2012} indicate the Laurentide ice sheet
may have formed a buttress to ice flow from the Western Cordillera, resulting
in a thicker Cordilleran ice sheet. However modelling alone will not elucidate
the nature and duration of the interaction between of the Cordilleran and
Laurentide ice sheets. For this, additional field constraints will be needed.

% ----------------------------------------------------------------------
\subsection{Past changes in precipitation}
% ----------------------------------------------------------------------

Another important source of error is the treatment of precipitation, to which
the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
studies, we base precipitation patterns on current values derived atmospheric
reanalyses. Although the distribution of precipitation has undoubtedly
been affected by past changes in global climate and the intermittent presence
of ice sheets and ice caps in the Northern Hemisphere, its changes are
difficult to assess. In the North American Cordillera, much of the continental
interior was likely dryer than today due to the presence of ice caps on the
Pacific Coast ranges, reinforcing the current topographic barrier. However,
some regions were probably rainier as a side effect of modified wind patterns.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown that different atmospheric reanalyses
--- circulation models using data assimilation techniques --- fail to agree
regarding the present-day pattern of precipitation over the modelling domain.
These discrepancies, which arise as a consequence of the highly mountainous
terrain, have dramatic effects on the ice sheet model response. Considering the
lack of observational data for modelling past climates, the increased
computational demand, and a decreased familiarity to processes at play, we must
expect that, for ice sheet simulations forced by different palaeo-climate
simulations, the uncertainty associated to the choice of atmospheric forcing
will be even higher. Perhaps alternative methods of coupling between ice sheet
evolution and precipitation changes need to be developed.

% ----------------------------------------------------------------------
\subsection{Geothermal heat flux patterns}
% ----------------------------------------------------------------------

Geothermal heat flux forcing, a major contributor to the modelled distribution
of en-glacial temperatures, and the location of basal melt areas, was assumed
constant over the modelling domain. However, the geology of the study area
consists of an amalgamation of former subduction zones, resulting in complex
geothermal patterns \citep{Blackwell.Richards.2004}. I am now at the stage to
start including this variability as a driver to the ice sheet model, which will
offer new opportunities to study its output in terms of ice flow patterns,
consistent basal freezing and basal melting patterns, regions prone for glacial
erosion, and regions characterised by glacial deposition.

I expect that inclusion of geothermal heat patterns will allow modelled basal
flow directions and distances to be compared with glacial lineations, which is
one of the main sources of geomorphological evidence used to reconstruct former
ice sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
instance, in our simulations, large portions of the Cordilleran ice sheet
remain cold-based throughout the deglaciation (\CCYC, Fig.~15). As cold-based
glaciers typically do not impact their beds significantly, this could indicate
that some of the lineation patterns we observe within these zones of consistent
cold-based deglaciation could be older features, formed during an earlier phase
of advance or retreat of the ice sheet. However, without a better control on
the geothermal heat flux, such analysis are perhaps too premature.

% ======================================================================
\section{Conclusions}
% ======================================================================

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, not been attained
before. Although mathematically uncertain, the coupled SIA-SSA approximation,
combined with massively parallel computing, allowed ice sheet modelling at
high spatial resolution and over multi-millenial time scales characteristic
of ice sheet growth and decay.
